---
title: "fahb simulation study"
format: html
editor: visual
---

## Introduction

This document records the design and analysis of a simulation study comparing methods for making progression decisions based on internal pilot recruitment data.

## Variables and paramerters

Using the syntax of the `fahb` package, we have the following design variables:

-   `m` - the number of sites to open over the course of the full trial.
-   `t` - the timing of the pilot interim analysis, expressed as a proportion of the expected time to recruit to target.
-   `N` - the target sample size which, when attained, will stop recruitment.

We also have some model parameters:

-   `beta` and `sigma` - mean and standard deviation parameters for the log-normal distribution of the site recruitment rates.
-   `lambda` - the rate describing the Poisson process of site opening.

Because we have a Bayesian model, the parameters to vary in a simulation study are actually the hyper-parameters describing the priors for the three substantive parameters above. We have:

-   `mu` and `nu` - mean and sd parameters of the normal prior on `beta`.
-   `rho` and `pi` - shape and rate parameters of the gamma prior on `sigma`.
-   `delta` and `epsilon` - shape and rate parameters of the gamma prior on `lambda`.

## Scenarios

We define scenarios by taking low or high values for:

-   The design parameters;
-   The expected prior values of the three substantive model parameters; and
-   An overall level of uncertainty in all three prior distributions.

For `m`, we will use 10 (low) or 30 (high).

For `t` we will use 0.2 (low) or 0.5 (high).

For `N` we will use 352 (low) or 788 (hgih). These correspond to total numbers of patients needed to detect an effect of 0.3 or 0.2 respectively, with a two-sided type I error rate of 0.05 and a power of 0.8.

For the expectation of `beta` we will use 6 (low) or 20 (high) patients per site per year.

For the expectation of `sigma` we will use 1 (low) or 3 (high).

For the expectation of `lambda` we will use 5 (low) and 15 (high) sites per year.

For each of these parameters and each of these settings, we want a version (defined through a choice of corresponding hyper-parameters) with low prior uncertainty and one with high prior uncertainty. First, `beta`:

```{r}
pred_params <- function(mean_rr_hps){
  # For a given parameter vector (mu, nu), calculate the expectation and sd
  # of a randomly sampled site recruitment rate
  mean <- exp(mean_rr_hps[1] + mean_rr_hps[2]^2/2) 
  sd <- sqrt(exp(mean_rr_hps[2]^2 + 2)*sqrt(exp(mean_rr_hps[2]^2) - 1))
  c(mean, sd)
}

# Low expectation, low variance
mean_rr_hps <- c(1.79, sqrt(2*(log(6) - 1.79)))
pred_params(mean_rr_hps)
mean_rr_hps
# 1.79, 0.05932064

# Low expectation, high variance
mean_rr_hps <- c(1.75, sqrt(2*(log(6) - 1.75)))
pred_params(mean_rr_hps)
mean_rr_hps
# 1.7500000 0.2889964

# High expectation, low variance
mean_rr_hps <- c(2.994, sqrt(2*(log(20) - 2.994)))
pred_params(mean_rr_hps)
mean_rr_hps
# 2.9940000 0.0588604

# Low expectation, high variance
mean_rr_hps <- c(2.954, sqrt(2*(log(20) - 2.954)))
pred_params(mean_rr_hps)
mean_rr_hps
# 2.9540000 0.2889023
```

Now `sigma`:

```{r}
library(extraDistr)

# Low expectation, low variance
sd_rr_hps <- c(100, 100/0.05)
sd_rr_hps[1]/sd_rr_hps[2]; sqrt(sd_rr_hps[1]/sd_rr_hps[2]^2)


# Low expectation, high variance
sd_rr_hps <- c(5, 5/0.05)
sd_rr_hps[1]/sd_rr_hps[2]; sqrt(sd_rr_hps[1]/sd_rr_hps[2]^2)


# High expectation (0.3), low variance
sd_rr_hps <- c(500, 500/0.3)
sd_rr_hps[1]/sd_rr_hps[2]; sqrt(sd_rr_hps[1]/sd_rr_hps[2]^2)


# High expectation, high variance
sd_rr_hps <- c(30, 30/0.3)
sd_rr_hps[1]/sd_rr_hps[2]; sqrt(sd_rr_hps[1]/sd_rr_hps[2]^2)
```

And finally, `lambda`:

```{r}
# Low expectation, low variance
so_hps <- c(200, 200/5)
so_hps[1]/so_hps[2]; sqrt(so_hps[1]/so_hps[2]^2)

# Low expectation, high variance
so_hps <- c(10, 10/5)
so_hps[1]/so_hps[2]; sqrt(so_hps[1]/so_hps[2]^2)

# High expectation, low variance
so_hps <- c(1800, 1800/15)
so_hps[1]/so_hps[2]; sqrt(so_hps[1]/so_hps[2]^2)

# High expectation, high variance
so_hps <- c(90, 90/15)
so_hps[1]/so_hps[2]; sqrt(so_hps[1]/so_hps[2]^2)
```

Now we put all these variations into a .csv file to be read in during the batch. Note we will replicate each scenario 10 times, running 10\^4 simulations in each component.

```{r}
m <- c(10, 30)
t <- c(0.2, 0.5)
N <- c(352,  788)

rec_rate <- c(1,2)
rec_v <- c(1,2)
setup_r <- c(1,2)

# First, make a matrix of values for the low prior uncertainty case
low_unc <- expand.grid(m, t, N, rec_rate, rec_v, setup_r)
low_unc$mu <- c(1.79, 2.994)[low_unc[,4]]
low_unc$nu <- c(0.0593, 0.0588)[low_unc[,4]]
low_unc$rho <- c(100, 500)[low_unc[,5]]
low_unc$pi <- c(2000, 1667)[low_unc[,5]]
low_unc$delta <- c(200, 1800)[low_unc[,6]]
low_unc$epsilon <- c(40, 120)[low_unc[,6]]

low_unc <- low_unc[,c(1:3, 7:12)]

# Similarly for high uncertainty
hi_unc <- expand.grid(m, t, N, rec_rate, rec_v, setup_r)
hi_unc$mu <- c(1.75, 2.954)[hi_unc[,4]]
hi_unc$nu <- c(0.2890, 0.2889)[hi_unc[,4]]
hi_unc$rho <- c(5, 30)[hi_unc[,5]]
hi_unc$pi <- c(100, 100)[hi_unc[,5]]
hi_unc$delta <- c(10, 90)[hi_unc[,6]]
hi_unc$epsilon <- c(2, 6)[hi_unc[,6]]

hi_unc <- hi_unc[,c(1:3, 7:12)]

scenarios <- rbind(low_unc, hi_unc)
scenarios <- cbind(1:nrow(scenarios), scenarios)

# Replicate each scenario 10 times to split across HPC
scenarios <- scenarios[rep(1:nrow(scenarios), each = 10),]
scenarios$part <- rep(1:10, 128)
scenarios <- scenarios[, c(1, 11, 2:10)]
names(scenarios)[1:5] <- c("id", "part", "m", "t", "N")

#write.csv(scenarios, file = paste0("test.csv"), row.names = FALSE)
```

To help visualise these priors we can plot them - for each of the three parameters we have a low and high mean and a low and high uncertainty:

```{r}
N <- 10^5
df <- data.frame(y = c(exp(rnorm(N, 1.79, 0.0593)),
                       exp(rnorm(N, 1.75,0.2890)),
                       exp(rnorm(N, 2.994, 0.0588)),
                       exp(rnorm(N, 2.954, 0.2889))),
                 val = rep(c("low", "high"), each = 2*N),
                 unc = rep(rep(c("low", "high"), each = N), 2))

p1 <- ggplot(df, aes(y, colour= unc)) + stat_density(alpha = 0) +
  xlim(c(0,30)) + xlab("Expected recruitment rate") + 
  theme_minimal()+
  scale_colour_discrete(name = "Uncertainty")
```

```{r}
df <- data.frame(y = c(rgamma(N, 100, 2000),
                       rgamma(N, 5, 100),
                       rgamma(N, 500, 1667),
                       rgamma(N, 30, 100)),
                 val = rep(c("low", "high"), each = 2*N),
                 unc = rep(rep(c("low", "high"), each = N), 2))

p2 <- ggplot(df, aes(y, colour= unc)) + stat_density(alpha = 0) +
  xlab("Standard deviation of recruitment rate") + 
  theme_minimal()+
  scale_colour_discrete(name = "Uncertainty")
```

```{r}
df <- data.frame(y = c(rgamma(N, 200, 40),
                       rgamma(N, 10, 2),
                       rgamma(N, 1800, 120),
                       rgamma(N, 90, 6)),
                 val = rep(c("low", "high"), each = 2*N),
                 unc = rep(rep(c("low", "high"), each = N), 2))

p3 <- ggplot(df, aes(y, colour= unc)) + stat_density(alpha = 0) +
  xlab("Site opening rate") + 
  theme_minimal()+
  scale_colour_discrete(name = "Uncertainty")

p1 / p2 / p3 + plot_layout(guides = 'collect') 

#ggsave("paper/priors.pdf", height=12, width=14, units="cm")
```

Each scenario will have a different expected time to recruit, which we can determine from the scenario values. Here we are assuming a simplified model of recruitment where recruitment rate increases linearly over time until we reach the maximum rate (`final_rate`) when all sites have opened (`all_open`); i.e. we will have `n = (final_rate / all_open) t^2 / 2` until time `all_open`, at which point we will have recruited `n_0 = (final_rate / all_open) all_open^2 / 2 = final_rate * all_open / 2` participants, and `n = n_0 + final_rate*(t - all_open)` thereafter:

```{r}
library(ggplot2)
final_rate <- 20
all_open <- 3
# While recruitment rate is increasing linearly
t1 <- seq(0, all_open, 0.01)
n1 <- (final_rate/all_open)*t1^2/2
# While recruitment rate is constant
t2 <- seq(all_open, 2*all_open, 0.01)
n2 <- final_rate*all_open/2 + final_rate*(t2 - all_open)

df <- data.frame(t = c(t1, t2),
                 n = c(n1, n2))

ggplot(df, aes(t, n)) + geom_line() +
  geom_vline(xintercept = all_open, linetype = 2) +
  xlab("Time") + ylab("Number recruited") +
  theme_minimal()
```

We therefore get the expected time to recruit the target of `N` depending on which segment of the recruitment curve we hit it:

```{r}
exp_rec_time <- function(sce){

  m <- sce[3]; N <- sce[5]
  mu <- sce[6]; nu <- sce[7]
  delta <- sce[10]; epsilon <- sce[11]
  
  # Expected rate of recruitment after all sites open
  final_rate <- exp(mu + nu^2/2)*m
  # Expected time until all sites open
  all_open <- m/(delta/epsilon)
  # Expected number recruited when all sites open
  n_0 <- final_rate*all_open/2
  if(n_0 >= N){
    # If we hit N before the max rate has been hit
    exp_time <- sqrt(2*N*all_open/final_rate)
  } else {
    exp_time <- all_open + (N - n_0)/final_rate
  }
  return(exp_time)
}

scenarios$exp_t <- apply(scenarios, 1, exp_rec_time)
```

We then use these expected recruitment times to define the underlying feasibility threshold for each scenario. Here, we classify a trial as truly feasible if its true time to recruit is less than 1.2 times its expected time to recruit:

```{r}
scenarios$thr <- 1.2*scenarios$exp_t
#write.csv(scenarios, file = paste0("paper/scenarios.csv"), row.names = FALSE)
```

## HPC

To keep the simulation study managable, we run it on HPC. Our `scenarios.csv` file has replicated each scenario ten times to help us batch the overall study into small chunks.

## Analysis

For every scenario, find the FPR/FNR curves for both the PC and Bayes approaches to analysis. This takes a bit of time (especially the MCO for the PC approach), so do it once and save the results.

```{r}
full_res <- NULL
for(i in 1:128){
  print(i)
  res_comp <- PC_OCs_curve(i)
  res_comp <- cbind(res_comp, Bayes_OCs_curve(i)[,2])
  res_comp <- cbind(i, res_comp)
  full_res <- rbind(full_res, res_comp)
}

# Note - we are getting a v. small proportions (max 0.5%) of NAs in the Bayesian
# mean predictions in a handful of scenarios

for(i in 1:128){
  full_res[full_res$sce == i, 7] <- Bayes_OCs_curve(i)[,2]
}

#saveRDS(full_res, "full_res.rds")
```

```{r}
full_res <- as.data.frame(readRDS("full_res.rds"))
names(full_res) <- c("sce", "fpr", "fnr_PC", "red_m", "red_n", "red_r", "fnr_B")
full_res$fnr_dif <- full_res$fnr_PC - full_res$fnr_B

ggplot(full_res, aes(x = fnr_dif)) + stat_density(alpha = 0.4) +
  xlab("Difference in optimal False Negative Rates") + 
  xlim(c(-0.025, 0.1)) +
  theme_minimal()

#ggsave("fnr_dif.pdf", height=9, width=14, units="cm")
```

Some numerical summaries:

```{r}
mean(full_res$fnr_dif)
quantile(full_res$fnr_dif, c(0.005, 0.995))
```

Look at redundancy in the optimal PC rules:

```{r}
colMeans(full_res[,4:6])
num_red <- full_res[,4] + full_res[,5] + full_res[,6]
mean(num_red == 0); mean(num_red == 1); mean(num_red == 2)
```

## Illustrative example

GUSTO data is saved under scenario 129. Analyse and add to the saved analysed data.

```{r}
full_res <- readRDS("full_res.rds")
for(i in 129){
  print(i)
  res_comp <- PC_OCs_curve(i)
  res_comp <- cbind(res_comp, Bayes_OCs_curve(i)[,2])
  res_comp <- as.data.frame(cbind(i, res_comp))
  names(res_comp) <- names(full_res)
  full_res <- rbind(full_res, res_comp)
}

# Note - we are getting a v. small proportions (max 0.5%) of NAs in the Bayesian
# mean predictions in a handful of scenarios

for(i in 129){
  full_res[full_res$sce == i, 7] <- Bayes_OCs_curve(i)[,2]
}

#saveRDS(full_res, "full_res.rds")
```

```{r}
full_res <- as.data.frame(readRDS("./paper/full_res.rds"))
full_res$fnr_dif <- full_res$fnr_PC - full_res$fnr_B

sub <- full_res[full_res$sce == 129,]

df <- data.frame(fpr = rep(sub$fpr, 2),
                 fnr = c(sub$fnr_PC, sub$fnr_B),
                 t = rep(c("PC", "Bayes"), each = nrow(sub)))

# Add the OCs of the progression criteria from the protocol - both
# the green crieteria and the non-red criteria.
df2 <- data.frame(fpr = c(0, 0.161),
                  fnr = c(0.961, 0.413),
                  t = c("PC", "PC"))

ggplot(df, aes(fpr, fnr, colour = t)) + geom_step() +
  geom_point(data= df2, colour = "black") +
  xlab("False Positive Rate") + ylab("False Negative Rate") +
  scale_colour_discrete(name = "Method") +
  theme_minimal()

#ggsave("ill_ex.pdf", height=9, width=14, units="cm")
```

Looking at a specific FPR:

```{r}
# Manually for now - need to re-run full analysis to keep the PC rule details
# 0.1994575937 0.334292258  1.04317670  7.33663681  5.5209017    0    0    0
df <- readRDS(paste0("./data/res_ext_", 129, "_full.rds"))
PC_OCs(c( 1.04317670,  7.33663681,  5.5209017), 3.6, df)
```

For GUSTO - target is 320 participants over 3 years from 20 sites.

Recruitment estimated at 13 per month once all sites are open - so 7.8 per site per year.

PC are at least 5 sites open and at least 30 participants recruited by 6 months.

If we assume a steady increase in recruitment rate as sites open, the above assumptions imply all sites will be open at 1.90 years into the study and so we get a site opening rate of 10.53 sites per year.

```{r}
beta_m <- 2; beta_s <- sqrt(2*(log(7.8) - beta_m))
beta_s
exp(beta_m + beta_s^2/2); sqrt(exp(beta_s^2 + 2)*sqrt(exp(beta_s^2) - 1))
hist(exp(rnorm(10^4, beta_m, beta_s)))

v_sh <- 30; v_r <- v_sh/0.3
v_r
v_sh/v_r; sqrt(v_sh/(v_r^2))
hist(rgamma(10^4, v_sh, rate = v_r))

beta <- rnorm(10^4, beta_m, beta_s)
sig <- rgamma(10^4, v_sh, rate = v_r)
lambdas <- exp(rnorm(10^4, beta, sig))
hist(lambdas)

setup_r_a <- 30; setup_r_b <- setup_r_a/10.53
setup_r_b
setup_r_a/setup_r_b; sqrt(setup_r_a/(setup_r_b^2))
hist(rgamma(10^4, setup_r_a, rate = setup_r_b))

gusto <- data.frame(id = 1,
                    part = 1:100,
                    m = 20,
                    int_t = 0.166,
                    target_n = 320,
                    beta_m = 2,
                    beta_s = 0.329,
                    v_sh = 30,
                    v_r = 100,
                    setup_r_a = 30,
                    setup_r_b = 2.85,
                    exp_t = 3,
                    thr = 1.2*3)

#write.csv(gusto, "R/gusto.csv", row.names = FALSE)
```

Plot the priors:

```{r}
df <- data.frame(beta = rnorm(10^5, beta_m, beta_s))
p1 <- ggplot(df, aes(beta)) + stat_density(alpha = 0.2, colour = "black") +
  xlab(expression(beta)) +
  theme_minimal()

df <- data.frame(sig = rgamma(10^5, v_sh, rate = v_r))
p2 <- ggplot(df, aes(sig)) + stat_density(alpha = 0.2, colour = "black") +
  xlab(expression(sigma)) +
  theme_minimal()

df <- data.frame(beta = rnorm(10^5, beta_m, beta_s),
                 sig = rgamma(10^5, v_sh, rate = v_r))
df$gamma <- exp(rnorm(10^5, df$beta, df$sig))
p3 <- ggplot(df, aes(gamma)) + stat_density(alpha = 0.2, colour = "black") +
  xlab(expression(gamma[i])) +
  xlim(c(0, 30)) +
  theme_minimal()

df <- data.frame(lambda = rgamma(10^5, setup_r_a, rate = setup_r_b))
p4 <- ggplot(df, aes(lambda)) + stat_density(alpha = 0.2, colour = "black") +
  xlab(expression(lambda)) +
  theme_minimal()

library(patchwork)
(p1 + p2) / (p3 + p4)

#ggsave("ill_priors.pdf", height=11, width=14, units="cm")
```

### Sensitivity

We would like to understand the consequences of not getting the standard PC thresholds quite right. To do so we can take an example efficient PC and then form a set of alternatives centered around that PC, and evaluate the operating characteristics of each one.

```{r}
#full_res <- as.data.frame(readRDS("./paper/full_res.rds"))
full_res$fnr_dif <- full_res$fnr_PC - full_res$fnr_B

sub <- full_res[full_res$sce == 129,]

df <- data.frame(fpr = rep(sub$fpr, 2),
                 fnr = c(sub$fnr_PC, sub$fnr_B),
                 t = rep(c("PC", "Bayes"), each = nrow(sub)))

ggplot(df, aes(fpr, fnr, colour = t)) + geom_step() +
  geom_point(data= df2, colour = "black") +
  xlab("False Positive Rate") + ylab("False Negative Rate") +
  scale_colour_discrete(name = "Method") +
  theme_minimal()

df2 <- as.data.frame(readRDS("./data/res_129_full.rds"))

base_rule <- c( 1.04317670,  7.33663681,  5.5209017)

rules <- matrix(base_rule * runif(3*1000, 0.5, 1.5), ncol = 3, byrow = TRUE)

sens_df <- t(apply(rules, 1, PC_OCs, thr=3.6, df=df2))

sens_df <- as.data.frame(sens_df)
names(sens_df) <- c("fpr", "fnr")

ggplot(df, aes(fpr, fnr, colour = t)) + geom_step() +
  geom_point(data= df2, colour = "black") +
  geom_point(data = sens_df, alpha = 0.9, colour = "black") +
  xlab("False Positive Rate") + ylab("False Negative Rate") +
  scale_colour_discrete(name = "Method") +
  theme_minimal()
```

## Extension

### External pilots

We can extend our approach easily to external pilots. The key changes are i) we reset everything after interim so that no sites are open and no patients have been recruited, and ii) we define the external pilot design by its numbers of participants and of sites. We can give a single method if we have a flag for resetting and if we define pilot sample sizes as caps and finish either when they are hit or when the interim analysis time point arrives - whichever comes first.

```{r}
scenarios <- read.csv("R/gusto.csv")[1,]

ext_design <- expand.grid(m_p = 3:10,
                          target_n_p = seq(30, 150, 10))

scenarios <- scenarios[rep(1, nrow(ext_design)),]
scenarios <- cbind(scenarios, ext_design)
scenarios$id <- 1:nrow(scenarios)

scenarios <- scenarios[rep(1:nrow(scenarios), each = 10),]
scenarios$part <- rep(1:10, 104)

# Change interim analysis time so that we always recruit to target in the pilot
scenarios$int_t <- 10

#write.csv(scenarios, file = "R/scenarios_ext.csv", row.names = FALSE)
```

```{r}
full_res <- NULL
for(i in 1:104){
  res_comp <- PC_OCs_curve(i)
  res_comp <- cbind(res_comp, Bayes_OCs_curve(i)[,2])
  res_comp <- cbind(i, res_comp)
  full_res <- rbind(full_res, res_comp)#
}

# Note - we are getting a v. small proportions (max 0.5%) of NAs in the Bayesian
# mean predictions in a handful of scenarios

#saveRDS(full_res, "full_res_ext.rds")

plot(full_res[full_res[,1] == 1, c(2,3)])
points(full_res[full_res[,1] == 1, c(2,7)], col="red")
```

```{r}
full_res <- readRDS("full_res_ext.rds")

ext_design <- expand.grid(m_p = 3:10,
                          target_n_p = seq(30, 150, 10))

# Collapse down to nominal FPR of 0.2
res <- full_res[full_res[,2] == 0.2,]

res <- as.data.frame(cbind(res[,c(2,3,7)], ext_design))
names(res)[2:3] <- c("PC_fnr", "Bayes_fnr")

ggplot(res, aes(m_p, target_n_p, fill=PC_fnr)) +
  geom_tile(color = "black") +
  scale_fill_gradient(low = "white", high = "darkgreen", name = "FNR") +
  xlab("Pilot sites") + ylab("Pilot participants") +
  theme_minimal()

#ggsave("ext_fnr.pdf", height=9, width=14, units="cm")

res$diff <- res$PC_fnr - res$Bayes_fnr
ggplot(res, aes(x = diff)) + stat_density(alpha = 0.4) +
  xlab("Difference in optimal False Negative Rates") + 
  xlim(c(-0.025, 0.05)) +
  theme_minimal()
```

### Bayesian SSD

On the face of it, the Bayesian approach is not feasible at the design stage. When it comes to choosing the rule we can argue that doesn't have to be a problem, because we don't need to pre-specify as a Bayesian. But a pre-specified rule can inform the sample size / the timing of the internal pilot analysis, so it would be useful.

We can address this by using the NP regression approach of Strong et al., which fits neatly here. Our methods work has shown that rules based on the standard PC items are very close to the OCs of the Bayesian rules, which tells us that an approximation built around those statistics should be very close too. So we can fit a GAM, getting an expected recruit time as a function of the simulated stats. We can then compare those predictions against the ones made by the full Bayesian method:

```{r}
library(mgcv)
df <- res_102_full
fit <- gam(rec_t ~ ti(m_p) + ti(n_p) + ti(r_p) + ti(m_p, n_p), data = df)
x <- (df$pred_t-predict.gam(fit))
q <- as.numeric(quantile(x, c(0.01, 0.99)))
y <- x[x>q[1] & x<q[2]]
hist(y)
```

This shows us that yes, the predictions from the GAM do tally well with the predictions made by the full Bayesian method. So it should give a good enough approximation for SSD. Doing that is easy - we proceed exactly as before when dealing with Bayesian rules, but now we use the predictions from the GAM rather than the full method.
